---
title: "Why Use OOP?"
aliases: why-oop
date: '2021-12-24'
abstract_short: ''
abstract: An example use case of S3 OOP in R.
math: yes
selected: yes
image:
  focal_point: ''
  preview_only: no
categories: Posts
tags:
- OOP
- S3
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(s3examples)
```

R is traditionally a functional programming language, meaning that when we code
in R, we tend to think in terms of functions. For instance, given a data set,
what function can I create to plot a certain variable? Object oriented
programming (OOP) introduces the concept of "classes." While classes are used
extensively in other programming languages such as Java, their use in R differs
from conventional uses. In part, this is because R has several unique OOP
systems which differ from one another in key ways. The three most popular
systems are S3, S4, and R6. While different R communities leverage different
systems, we will focus on S3, which is favored by the folks over at
[RStudio](https://www.tidyverse.org/), the
[Tidyverse](https://www.tidyverse.org/), and other related organizations. For
more information, see the [additional reading](#additional-reading) section
below.

## An Example Use Case

Consider the case where you want to create a function that works for a specific
type of data format. Perhaps there you are interested in a specific column and
want to ensure your data has this column present. For instance, let us take a
look at the following dataset.

```{r example data, echo = FALSE}
ref_file <- miplicorn::miplicorn_example("reference_AA_table.csv")
alt_file <- miplicorn::miplicorn_example("alternate_AA_table.csv")
cov_file <- miplicorn::miplicorn_example("coverage_AA_table.csv")

data <- miplicorn::read_tbl_ref_alt_cov(
  ref_file,
  alt_file,
  cov_file,
  gene == "atp6" | gene == "crt"
)

data
```

Let us say we want to create a function `mutation_prevalence` which determines
the prevalence of each mutation in the dataset. Moreover, we would like to plot
this prevalence.

```{r mutation prev}
prevalence <- mutation_prevalence(data, 5)
prevalence
```

While creating this function may require some thought, it makes intuitive sense
that the result will contain four columns as seen above. Two of these columns,
`n_total` and `n_mutant`, are used to compute the final column, `prevalence`.
Therefore, if we wanted to visualize our data, it is not very important for us
to consider the two columns `n_total` and `n_mutant`. Rather, we care about
the `mutation_name` and the `prevalence`.

If we were to create an function, `plot_prevalence`, the easiest way to code it
would be to give the function a `data` argument:

```{r basic fn, eval = FALSE}
plot_prevalence <- function(data) {
  # code for plotting goes here
}
```

Notice that in the given format, we could feed any dataset into our function.
In other words, there are no checks to make sure we are feeding in the output
of `mutation_prevalence()`. How could we address this potential problem?

### Potential Solutions

The most obvious solution would be to check that some key columns exist:

```{r key columns, eval = FALSE}
plot_prevalence <- function(data) {
  if (!"mutation_name" %in% colnames(data)) {
    stop("Missing key column!", call. = FALSE)
  }

  # code for plotting goes here
}
```

Howerver, another strategy would be to use classes! If we assigned a class, say
`mut_prev`, to the output of `mutation_prevalence()`, we could easily check if
the input is of type `mut_prev`:

```{r class check, eval = FALSE}
plot_prevalence <- function(data) {
  if (!inherits(data, "mut_prev")) {
    stop("Wrong class input!", call. = FALSE)
  }

  # code for plotting goes here
}
```

## But Wait, There's More!

Now, using classes to solve this issue might seem like overkill.
After all, why create an entire new class when you can just have an `if()`
statement? The real power of object-oriented programming is the ability to use
**polymorphism**[^1]. What we mean by this is the
ability to use the same function for many types of input. An example of this
behavior is the base R function `print()` which behaves differently depending on
the class of the input.

```{r print behavior}
# A vector
print(c(1, 2))

# A tibble
print(tibble::as_tibble(c(1, 2)))
```

Using OOP, developers can customize how certain functions interact with certain
objects. Revisiting our `mut_prev` class, we could even change the way the table
is printed! Another common example is plotting a dataset. Say we have developed
a package that introduces three different classes of datasets. We can then
create a plot method for each type of class.

```{r print methods, eval = FALSE}
plot.class1 <- function(x) {
  # plotting code
}

plot.class2 <- function(x) {
  # plotting code
}

plot.class3 <- function(x) {
  # plotting code
}
```

We can witness this in action with our class `mut_prev`. When we call `plot()`
a custom plot specific to only our class is generated!

```{r plot class}
plot(prevalence)
```

[^1]: The term polymorphism has been taken from [Advanced
R](https://adv-r.hadley.nz/oo.html#oop-systems).

## Additional Reading

There is a lot more to the world of OOP and many important packages leverage
various OOP techniques. In fact, as we mentioned earlier, there are even
multiple OOP systems in R. For example, the
[Tidyverse](https://www.tidyverse.org/) is built using the S3 system whereas the
[Bioconductor project](https://www.bioconductor.org/) uses S4. Of all the
available systems, S3 is regarded as the easiest system to learn.

To learn more about OOP, I would highly recommend reading the OOP chapter of
[Advanced R](https://adv-r.hadley.nz/oo.html) and visiting some of the articles
in the [`{vctrs}` package](https://vctrs.r-lib.org/articles/s3-vector.html). The
functions used for this post can be found in my [`{s3examples}`
package](https://github.com/arisp99/s3examples) where I have written a simple S3
subclass of the `tibble()` class.
